Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding optimization Choices

### 2025-03-03 jso.qmd
proofread before Sec 3.2

---
title: Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: Guangdong Provincial/Zhuhai Key Laboratory of Interdisciplinary Research and Application for Data Science, BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Wai Yin Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
abstract: |
  The projection pursuit (PP) guided tour interactively optimizes a criterion function known as the PP index, to explore high-dimensional data by revealing interesting projections. Optimization of some PP indexes can be non-trivial, if they are non-smooth functions, or the optimum has a small "squint angle", detectable only from close proximity. Here, measures for calculating the smoothness and squintability properties of the PP index are defined. These are used to investigate the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimizer (JSO), for optimizing PP indexes. The performance of JSO for visualizing data is evaluated across various hyper-parameter settings and compared with existing optimisers for the guided tour. The JSO algorithm has been implemented in the R package, `tourr`, and functions to calculate smoothness and squintability measures are implemented in the `ferrn` package.
keywords: 
  - projection pursuit
  - jellyfish search optimiser (JSO)
  - optimisation
  - grand tour
  - high-dimensional data
  - exploratory data analysis
date: last-modified
bibliography: bibliography.bib
format:
  tandf-pdf:
    keep-tex: true
    fontsize: 12pt
    linestretch: 2
editor: 
  markdown: 
    wrap: 72
crossref: 
  eq-prefix: ""
editor_options: 
  chunk_output_type: console
header-includes:
   - \usepackage{algorithm}
   - \usepackage{float} 
   - \usepackage{amsmath}
   - \usepackage{amsthm}
   - \usepackage{amssymb}
   - \theoremstyle{plain}
   - \newtheorem{defn}{\protect\definitionname}
   - \newtheorem{prop}{\protect\propositionname}
   - \providecommand{\definitionname}{Definition}
   - \providecommand{\propositionname}{Proposition}
nocite: |
   @DLMF
---

@@

Projection pursuit (PP) [@kr69; @FT74; @huber85] is a dimension
reduction technique aimed at identifying informative linear projections
of data. This is useful for exploring high-dimensional data, and
creating plots of the data that reveal the main features to use for
publication. The method involves optimizing an objective function known
as the PP index [e.g., @hall1989polynomial; @cook1993projection;
@lee2010projection; @Loperfido2018; @Loperfido2020], which defines the
criterion for what constitutes interesting or informative projections.
Let $X \in \mathbb{R}^{n\times p}$ be the data matrix,
$A \in\mathbb{R}^{p \times d}$ be an orthonormal matrix, where $A$
belongs to the Stiefel manifold $\mathcal{A} = V_d(\mathbb{R}^p)$. The
projection $Y = XA$ is a linear transformation that maps data from a
$p$-dimensional space into a $d$-dimensional space. The index function
$f(XA): \mathbb{R}^{n \times d} \to \mathbb{R}$ is a scalar function
that measures an interesting aspect of the projected data, such as
deviation from normality, presence of clusters, or non-linear structure.
For a fixed sample of data, PP finds the orthonormal basis $A$ that
maximizes the index value of the projection, $Y = XA$:

@@

It is interesting to note that when using PP visually, one cares less
about $A$ than the plane described by $A$, because the orientation in
the plane is irrelevant. The space of planes belongs to a Grassmann
manifold. This is usually how the projection pursuit guided tour (PPGT)
[@cook1995grand] operates, when using geodesic interpolation between
starting and target planes. It interpolates plane to plane, removing
irrelevant within plane spin, and is agnostic to the basis ($A$) used to
define the plane. Thus, indexes which are used for the PPGT should be
rotationally invariant.

Index functions are quite varied in form, partially depending on the
data that is being projected. @fig-example-functions shows two examples.
Huber plots [@huberplot] of 2D data sets are in (a) and (c), showing the
PP index values for all 1D projections of the 2D data in polar
coordinates, which reveals the form of these functions. The dashed
circle is a baseline set at the average value, and the straight line
marks the optimal projection. Plots (b) and (d) show the respective best
projections of the data as histograms. Indexes like the holes, central
mass and skewness [@cook1993projection] are generally smooth for most
data sets, but capture only large patterns. Many indexes are noisy and
non-convex, requiring an effective and efficient optimisation procedure
to explore the data landscape and achieve a globally optimal viewpoint
of the data. The skewness index computed for trimodal data, in (a), is
smooth with a large squint angle but has three modes, and thus is not
convex. The binned normality index [a simple version of a non-normality
index as described in @huber85] computed on the famous RANDU data, in
(c), is noisier and has a very small squint angle. The discreteness
cannot be seen unless the optimiser is very close to the optimal
projection.

@@

Optimization of PP is often discussed when new indexes are proposed
[@posse95; @marie-sainte2010; @grochowski2011]. @cook1995grand tied the
optimisation more closely to the index, when they introduced the PPGT,
which monitors the optimisation visually so that the user can see the
projected data leading in and out of the optimum. An implementation is
available in the `tourr` package [@tourr] in R [@R]. @RJ-2021-105
illustrated how to diagnose optimisation processes, particularly
focusing on the guided tour, and revealed a need for improved
optimisation. While improving the quality of the optimisation solutions
in the tour is essential, it is also important to be able to view the
data projections as the optimisation progresses. Integrating the guided
tour with a global optimisation algorithm that is efficient in finding
the global optimal and enables viewing of the projected data during the
exploration process is a goal.

Here, the potential for a Jellyfish Search Optimizer (JSO) [see
@chou_novel_2021, and @rajwar_exhaustive_2023] for the PPGT is explored.
JSO, inspired by the search behavior of jellyfish in the ocean, is a
swarm-based metaheuristic designed to solve global optimisation
problems. Compared to traditional methods, JSO has demonstrated stronger
search ability and faster convergence, and requires fewer tuning
parameters. These practicalities make JSO a promising candidate for
enhancing PP optimisation.

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a
global optimum, for various PP indexes that may have differing
complexities. To observe the performance of JSO with different types of
PP indexes, metrics are introduced to capture specific properties of the
index including squintability [based on @barnett1981interpreting's
squint angle] and smoothness. Here, we mathematically define metrics for
squintability and smoothness, which is a new contribution for PP
research. A series of simulation experiments using various datasets and
PP indexes are conducted to assess JSO's behavior and its sensitivity
to hyper-parameter choices (number of jellyfish and maximum number of
tries). The relationship between the JSO performance, hyper-parameter
choices and properties of PP indexes (smoothness and squintability) is
analysed to provide guidance on selecting optimisers for practitioners
using projection pursuit. Additionally, this work should guide the
design of new PP indexes and facilitate better optimization for PP.

The paper is structured as follows. @sec-background introduces the
background of the PPGT, reviews existing optimisers and index functions
in the literature. @sec-PP-properties introduces the metrics that
measure two properties of PP indexes, smoothness and squintability, and
@sec-JSO describes the new JSO to be used for the PPGT. @sec-sim-deets
outlines two simulation experiments to assess JSO's performance: one
comparing JSO's performance improvements relative to an existing
optimiser, Creeping Random Search (CRS), and the other studying the
impact of PP index properties on optimisation performance, and
@sec-sim-res presents the results. @sec-discussion discusses the
implementation of JSO in the `tourr` package and the PP property
calculation in the `ferrn` package. @sec-conclusion summarises the work
and provides suggestions for future directions.

# Projection pursuit, tours, index functions and optimisation {#sec-background}

A tour on high-dimensional data is constructed by geodesically
interpolating between pairs of planes. Any plane is described by an
orthonormal basis, $A_t$, where $t$ represents time in the sequence. The
term "geodesic" refers to maintaining the orthonormality constraint so
that each view shown is correctly a projection of the data. The PP
guided tour operates by geodesically interpolating to target planes
(projections) which have high PP index values, as provided by the
optimiser. The geodesic interpolation means that the viewer sees a
continuous sequence of projections of the data, so they can watch
patterns of interest forming as the function is optimised. There are
five optimisation methods implemented in the `tourr` package:

-   a pseudo-derivative, that searches locally for the best direction,
    based on differencing the index values for very close projections.
-   a brute-force optimisation (CRS).
-   a modified brute force algorithm described in @posse95.
-   an essentially simulated annealing [@Bertsimas93] where the search
    space is reduced during the optimisation.
-   a very localized search, to take tiny steps to get closer to the
    local maximum.

```{=html}
<!-- 
-   `search_geodesic()`: provides a pseudo-derivative optimisation. It
    searches locally for the best direction, based on differencing the
    index values for very close projections. Then it follows the
    direction along the geodesic path between planes, stopping when the
    next index value fails to increase.
-   `search_better()`: also known as Creeping Random Search (CRS), is a
    brute-force optimisation searching randomly for projections with
    higher index values.
-   `search_better_random()`: is essentially simulated annealing
    [@Bertsimas93] where the search space is reduced as the optimisation
    progresses.
-   `search_posse()`: implements the algorithm described in @posse95.
-   `search_polish()`: is a very localised search, to take tiny steps to
    get closer to the local maximum.
-->
```
There are numerous PP index functions available: introduced in @huber85,
@cook1993projection, @lee2005projection, @lee2010projection, @Grimm2016,
@Laa:2020wkm. Most are relatively simply defined, for any projection
dimension, and implemented because they are relatively easy to optimist.
A goal is to develop PP indexes based on scagnostics [@scag, @WW08], but
the blockage is their optimisation as these tend to be noisy, with
potentially small squint angles.

An initial investigation of PP indexes, and the potential for
scagnostics is described in @laa_using_2020. To be useful here an
optimiser needs to be able to handle index functions that are possibly
not very smooth. In addition, because data structures might be
relatively fine, the optimiser needs to be able to find maxima that
occur with a small squint angle, that can only be seen from very close
by. One last aspect that is useful is for an optimiser to return local
maxima in addition to the global one because data can contain many
different and interesting features.

@@

@laa_using_2020 has proposed five criteria for assessing projection
pursuit indexes (smoothness, squintability, flexibility, rotation
invariance, and speed). Since not all index properties affect the
optimisation process, the focus here is on the first two properties,
*smoothness* (@sec-smoothness) and *squintability* (@sec-squintability),
for which metrics are proposed to quantify them.

@@

This subsection proposes a metric for the smoothness of a projection
pursuit index.

@@

```{=tex}
\begin{defn}[Sobolev space]\label{def:sobolev_space}
The Sobolev space $W^{k,p}(\mathbb{R})$ for $1\leq p\leq \infty$ is the set of all functions $f$ in $L^p(\mathbb{R})$ for which all weak derivatives $f^{(\ell)}$ of order $\ell\leq k$ exist and have a finite $L^p$ norm.
\end{defn}
```
The Sobolev index $k$ in Definition \ref{def:sobolev_space} can be used
to characterize the smoothness of a function: if $f\in W^{k,p}$, then
the higher $k$, the smoother $f$. While this Sobolev index $k$ is a
useful measure of smoothness, it can be difficult to compute or even
estimate in practice.

To obtain a computable estimator of the smoothness of the index function
$f$, we propose an approach based on random fields. If a PP index
function $f$ is evaluated at some random bases, as is done at the
initialization stage of JSO, then these random index values can be
interpreted as a random field, indexed by a space parameter, namely the
random projection basis. This analogy suggests to use this random
training sample to fit a spatial model. We propose to use a Gaussian
process equipped with a Matérn covariance function, due to the
connections between this model and Sobolev spaces, see for example
@porcu2024matern.

@@

In view of these results, the parameter $\nu$ is suggested as a measure
of the smoothness of the PP index function by fitting a Gaussian process
prior with Matérn covariance on a dataset generated by random
evaluations of the index function, as done at the initialization stage
of the jellyfish search optimization. There exist several R packages,
such as `GpGp` [@guinness2021gpgp] or `ExaGeoStatR` [@abdulah2023large],
to fit the hyper-parameters of a GP covariance function on data, which
is usually done by maximum likelihood estimation. In this project, the
`GpGp` package is used.

```{=tex}
\begin{defn}
Let $\mathbf{A}=[A_1, \ldots, A_N] \in (\mathbb{R}^{p \times d})^N$ be d-dimensional projection bases, let $\mathbf{y}=[f(XA_1),\ldots,f(XA_N)]$ be the corresponding PP index values, and let $\mathbf{K}=[K_\theta(A_{i}-A_{j})]_{1\leq i,j\leq N}\in\mathbb{R}^{N\times N}$ be the Matérn covariance matrix evaluated at the input bases, where the vector $\theta$ contains all the parameters of the multivariate Matérn covariance function $K$ (smoothness, outputscale, lengthscales). The log-likelihood of the parameters $\theta$ is defined by 
\begin{equation}
\mathcal{L}(\theta)=\log p(\mathbf{y}\left|\mathbf{A},\theta\right.)=-\frac{1}{2}\mathbf{y}^{\top}(\mathbf{K}+\sigma^{2}\mathbf{I})^{-1}\mathbf{y}-\frac{1}{2}\mathrm{\log}(\det(\mathbf{K}+\sigma^{2}\mathbf{I}))-\frac{N}{2}\log(2\pi)\, \label{eq:gp_log_likelihood}
\end{equation}
where the nugget parameter $\sigma$ is the standard deviation of the intrinsic noise of the Gaussian process.
The optimal parameters (including smoothness) are obtained by maximum log-likelihood
\begin{equation}
\theta^* = \underset{\theta}{\max}\mathcal{L}(\theta)
\end{equation}
The resulting optimal smoothness parameter $\nu$ is chosen as our smoothness metric.
\end{defn}
```
The value of the optimal smoothness parameter $\nu>0$ can be naturally
interpreted as follows: the higher $\nu$, the smoother the index
function.

@@

This section defines the projection distance and introduces the
squintability metric, followed by a description of two numerical
approaches for computing squintability.

```{=tex}
\begin{defn}[projection distance]\label{def:proj-dist}
Let $A \in \mathbb{R}^{p \times d}$ be a $d$-dimensional orthonormal matrix, and let $A^*$ be the optimal matrix that achieves the maximum index value for a given data. The projection distance between $A$ 
and $A^*$, $r(A, A^*)$, is defined by
$r(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\,\rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}
\begin{defn}[squint angle]\label{def:squint-angle}

Let $A$ and $B$ be two $d$-dimensional orthonormal matrices in $\mathbb{R}^p$. The squint angle $\theta$ between the subspace spanned by $A$ and $B$ is defined as the smallest principal angle between these subspaces: $\theta = \min_{i \in \{1, \cdots, d\}} \arccos(\tau_i)$, where $\tau_i$ are the singular values of the matrix $M = A^T B$ obtained from its singular value decomposition.
\end{defn}
```
Squintability can be defined based on how the index value $f(XA)$ changes with
respect to the projection distance $r(A, A^*)$, over the course of the
JSO. Suppose that the optimization starts with an arbitrary candidate
matrix $A_0$ at a distance $r_0:=r(A_0, A^*)$ from the optimal one. A possible way to define squintability is as follows:

@@

From @barnett1981interpreting and @laa_using_2020, a large squint angle
implies that the objective function value is close to optimal even when
the perfect view to see the structure is far away. A small squint angle
means that the PP index value improves substantially only when the
perfect view is close by. As such, low squintability implies rapid
improvement in the index value when near the perfect view. For PP, a
small squint angle is considered to be undesirable because it means that
the optimiser needs to be very close to be able to "see" the optimum.
Thus, it could be difficult for the optimiser to find the optimum.
<!--The
mathematical formulation of this intuition is proposed below:-->

@@

# The jellyfish optimiser {#sec-JSO}

@@

To solve the optimisation problem embedded in the PP guided tour, a
starting projection, an index function, the number of jellyfish, and the
maximum number of trials (tries) are provided as input. Then, the
current projection is evaluated by the index function. The projection is
then moved in a direction determined by a random factor, influenced by
how far along we are in the optimisation process. Occasionally,
completely new directions may be taken like a jellyfish might with ocean
currents. A new projection is accepted if it is an improvement compared
to the current one, rejected otherwise. This process continues and
iteratively improves the projection, until the pre-specified maximum
number of trials is reached.

@@

**Output**: `optimised_projection`

@@

**return** the set of projections with the updated `current_best` as the
`optimised_projection`
:::

@@

# Assessing the optimisers {#sec-sim-deets}

This section explains the details of two simulation studies: (1) comparison between the existing Creeping Random Search (CRS) [@RJ-2021-105; @laa_using_2020] and JSO, and (2) examining the factors affecting the performance of JSO.

@@

The CRS is the main optimisation routine currently used for the guided tour. Here the two optimisers are compared on the task of finding the pipe using the `holes` index in data with dimensions, $d = 6, 8, 10, 12$. Fifty simulated data sets are used for each dimension. JSO uses 100 jellyfish with a maximum of 100 tries, while the CRS allows a maximum of 1000 tries at each iteration before the algorithm terminates. These choices enable fair comparison between CRS and JSO, that conforms to how they are used in practice.

>>>>

---
title: Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding optimization Choices
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: Guangdong Provincial/Zhuhai Key Laboratory of Interdisciplinary Research and Application for Data Science, BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Wai Yin Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
abstract: |
  The projection pursuit (PP) guided tour optimizes a criterion function, known as the PP index, to gradually reveal projections of interest from high-dimensional data through animation. Optimization of some PP indexes can be non-trivial, if they are non-smooth functions, or when the optimum has a small "squint angle", detectable only from close proximity. Here, measures for calculating the smoothness and squintability properties of the PP index are defined. These are used to investigate the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimizer (JSO), for optimizing PP indexes. The performance of JSO in detecting the target pattern (pipe shape) is compared with existing optimizers in PP. Additionally, JSO's performance on detecting the sine-wave shape is evaluated using different PP indexes (hence different smoothness and squintability) across various data dimensions (d = 4, 6, 8, 10, 12) and JSO hyper-parameters. The JSO algorithm has been implemented in the R package, `tourr`, and functions to calculate smoothness and squintability measures are implemented in the `ferrn` package.
keywords: 
  - Projection Pursuit Guided Tour (PPGT)
  - Jellyfish Search Optimizer (JSO)
date: last-modified
bibliography: bibliography.bib
format:
  tandf-pdf:
    keep-tex: true
    fontsize: 12pt
    linestretch: 2
editor: 
  markdown: 
    wrap: 72
crossref: 
  eq-prefix: ""
editor_options: 
  chunk_output_type: console
header-includes:
   - \usepackage{algorithm}
   - \usepackage{float} 
   - \usepackage{amsmath}
   - \usepackage{amsthm}
   - \usepackage{amssymb}
   - \theoremstyle{plain}
   - \newtheorem{defn}{\protect\definitionname}
   - \newtheorem{prop}{\protect\propositionname}
   - \providecommand{\definitionname}{Definition}
   - \providecommand{\propositionname}{Proposition}
nocite: |
   @DLMF
---

@@

Projection pursuit (PP) [@kr69; @FT74; @huber85] is a dimension
reduction technique aimed at identifying informative linear projections
of data. This is useful for exploring high-dimensional data, and
creating plots of the data that reveal the main features to use for
publication. The method involves optimizing an objective function known
as the PP index [e.g., @hall1989polynomial; @cook1993projection;
@lee2010projection; @Loperfido2018; @Loperfido2020], which defines the
criterion for what constitutes interesting or informative projections.
Let $X \in \mathbb{R}^{n\times p}$ be the data matrix,
$A \in\mathbb{R}^{p \times d}$ be an orthonormal matrix in the Stiefel 
manifold $\mathcal{A} = V_d(\mathbb{R}^p)$. The
projection $Y = XA$ is a linear transformation of the $p$-dimensional data to
a $d$-dimensional space. The index function
$f(XA): \mathbb{R}^{n \times d} \to \mathbb{R}$ defines a statistics to measure
an pattern of interest in the projected data, such as
deviation from normality, presence of clusters, or non-linear structure.
For a fixed sample of data, PP finds the orthonormal matrix $A$ (also called 
*projection basis* in PP literature) that maximizes the index value of the 
projection:

@@

where $I_d$ is the identity matrix of dimension $d$.  
It is interesting to note that when using PP visually, one cares less
about $A$ than the plane described by $A$, because the orientation in
the plane is irrelevant. The space of planes belongs to a Grassmann
manifold. This is usually how the projection pursuit guided tour (PPGT)
[@cook1995grand] operates, when using geodesic interpolation between
starting and target planes. It interpolates plane to plane, removing
irrelevant within plane spin, and is agnostic to the basis ($A$) used to
define the plane. Thus, indexes which are used for the PPGT should be
rotationally invariant.

Index functions are quite varied in form, partially depending on the
data that is being projected. @fig-example-functions shows two examples.
Huber plots [@huberplot] of 2D data sets are in (a) and (c), showing the
PP index values for all 1D projections of the 2D data in polar
coordinates, which reveals the form of these functions. The dashed
circle is a baseline set at the average value, and the straight line
marks the optimal projection. Plots (b) and (d) show the respective best
projections of the data as histograms. Indexes like the holes, central
mass and skewness [@cook1993projection] are generally smooth for most
data sets, but capture only large patterns. Many indexes are noisy and
non-convex, requiring an effective and efficient optimization procedure
to explore the data landscape and achieve a globally optimal viewpoint
of the data. The skewness index computed for trimodal data, in (a), is
smooth with a large squint angle but has three modes, and thus is not
convex. The binned normality index [a simple version of a non-normality
index as described in @huber85] computed on the famous RANDU data, in
(c), is noisier and has a very small squint angle. The discreteness
cannot be seen unless the optimizer is very close to the optimal
projection.

@@

Optimization of PP is often discussed when new indexes are proposed
[@posse95; @marie-sainte2010; @grochowski2011]. @cook1995grand tied the
optimization more closely to the index, when they introduced the PPGT,
which monitors the optimization visually so that the user can see the
projected data leading in and out of the optimum. An implementation is
available in the `tourr` package [@tourr] in R [@R]. @RJ-2021-105
illustrated how to diagnose optimization processes, particularly
focusing on the guided tour, and revealed a need for improved
optimization. While improving the quality of the optimization solutions
in the tour is essential, it is also important to be able to view the
data projections as the optimization progresses. Integrating the guided
tour with a global optimization algorithm that is efficient in finding
the global optimal and enables viewing of the projected data during the
exploration process is a goal.

In this work, the potential for a Jellyfish Search Optimizer (JSO) [
@chou_novel_2021; @rajwar_exhaustive_2023] for the PPGT is explored.
The JSO, inspired by the search behavior of jellyfish in the ocean, is a
swarm-based metaheuristic designed to solve global optimization
problems. Compared to other metaheuristic methods, JSO has demonstrated stronger
search ability and faster convergence, and requires fewer tuning
parameters. These practicalities make JSO a promising candidate for
enhancing PP optimization.

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimization for the guided tour. It is of
interest to assess how quickly and closely the optimizer reaches a
global optimum, for various PP indexes that may have differing
complexities. To observe the performance of JSO with different types of
PP indexes, metrics are introduced to capture specific properties of the
index including squintability [based on @barnett1981interpreting's
squint angle] and smoothness. We mathematically define metrics for
squintability and smoothness, which is a new contribution for PP
research. A series of simulation experiments are conducted using JSO to detect 
different target patterns (pipe and sine-wave) with different PP indexes 
(holes, MIC, TIC, dcor, loess, splines, skinny, and stringy), as well as JSO's 
hyper-parameter choices (number of jellyfish and maximum number of
tries). This work should facilitate better optimization for PP and
guide the choice of optimizer when designing new PP indexes.

The paper is structured as follows. @sec-background introduces the
background of the PPGT, reviews existing optimizers and index functions
in the literature. @sec-PP-properties introduces the metrics that
measure two properties of PP indexes, smoothness and squintability. 
@sec-JSO describes the new JSO for PP. @sec-sim-deets
outlines two simulation experiments to assess JSO's performance: one
comparing JSO's performance improvements relative to an existing
optimizer, Creeping Random Search (CRS), and the other studying the
impact of PP index properties, controlling for data dimension and JSO 
hyper-parameters, on optimization performance, and
@sec-sim-res presents the results. @sec-discussion discusses the
implementation of JSO in the `tourr` package and the PP property
calculation in the `ferrn` package. @sec-conclusion summarizes the work
and provides suggestions for future directions.

# Projection pursuit, tours, index functions and optimization {#sec-background}

A tour on high-dimensional data is constructed by geodesically
interpolating between pairs of planes. Any plane is described by an
orthonormal basis, $A_t$, where $t$ represents time in the sequence. The
term "geodesic" refers to maintaining the orthonormality constraint so
that each view shown is correctly a projection of the data. The PP
guided tour operates by geodesically interpolating to target planes
(projections) which have high PP index values, as provided by the
optimizer. The geodesic interpolation means that the viewer sees a
continuous sequence of projections of the data, so they can watch
patterns of interest forming as the function is optimized. There are
five optimization methods implemented in the `tourr` package:

-   a pseudo-derivative, that searches locally for the best direction,
    based on differencing the index values for very close projections.
-   a brute-force optimization (CRS).
-   a modified brute force algorithm described in @posse95.
-   an essentially simulated annealing [@Bertsimas93] where the search
    space is reduced during the optimization.
-   a very localized search, to take tiny steps to get closer to the
    local maximum.

```{=html}
<!-- 
-   `search_geodesic()`: provides a pseudo-derivative optimization. It
    searches locally for the best direction, based on differencing the
    index values for very close projections. Then it follows the
    direction along the geodesic path between planes, stopping when the
    next index value fails to increase.
-   `search_better()`: also known as Creeping Random Search (CRS), is a
    brute-force optimization searching randomly for projections with
    higher index values.
-   `search_better_random()`: is essentially simulated annealing
    [@Bertsimas93] where the search space is reduced as the optimization
    progresses.
-   `search_posse()`: implements the algorithm described in @posse95.
-   `search_polish()`: is a very localized search, to take tiny steps to
    get closer to the local maximum.
-->
```
There are numerous PP index functions available: introduced in @huber85,
@cook1993projection, @lee2005projection, @lee2010projection, @Grimm2016,
@Laa:2020wkm. Most are relatively simply defined, for any projection
dimension, and implemented because they are relatively easy to optimize.
A goal is to develop PP indexes based on scagnostics [@scag; @WW08], but
the blockage is their optimization as these tend to be noisy, with
potentially small squint angles.

An initial investigation of PP indexes, and the potential for
scagnostics is described in @laa_using_2020. To be useful here an
optimizer needs to be able to handle index functions that are possibly
not very smooth. In addition, because the target structure in the data might be
relatively fine, the optimizer needs to be able to find maxima that
occur with a small squint angle, that can only be seen from very close
by. One last aspect that is useful is for an optimizer to return local
maxima in addition to the global one because data can contain many
different and interesting features.

@@

@laa_using_2020 has proposed five criteria for assessing projection
pursuit indexes (smoothness, squintability, flexibility, rotation
invariance, and speed). Since not all index properties affect the
optimization process, the focus of this work is on the first two properties,
*smoothness* (@sec-smoothness) and *squintability* (@sec-squintability),
for which metrics are proposed to quantify them.

@@

@@

```{=tex}
\begin{defn}[Sobolev space]\label{def:sobolev_space}
The Sobolev space $W^{k,p}(\mathbb{R})$ for $1\leq p\leq \infty$ is the set of all functions $f$ in $L^p(\mathbb{R})$ for which all weak derivatives $f^{(\ell)}$ of order $\ell\leq k$ exist and have a finite $L^p$ norm.
\end{defn}
```

The Sobolev index $k$ in Definition \ref{def:sobolev_space} can be used
to characterize the smoothness of a function: if $f\in W^{k,p}$, then
the higher $k$, the smoother $f$. While this Sobolev index $k$ is a
useful measure of smoothness, it can be difficult to compute or even
estimate in practice. For example, the loess and splines indexes are not
continuous due to the use of the $\max$ operator in their definitions. 
Scagnostic indexes, such as skinny and stringy, are defined based on
graph elements (area, diameter, perimeter, and length) of alph hull or 
minimal spanning tree).

To obtain a computable estimator of the smoothness of the index function
$f$, we propose an approach based on random fields. If a PP index
function $f$ is evaluated at some random bases, then these random index values 
can be interpreted as a random field, indexed by a space parameter, namely the
random projection basis. This analogy suggests to use this random
training sample to fit a spatial model. We propose to use a Gaussian
process equipped with a Matérn covariance function, due to the
connections between this model and Sobolev spaces, see for example
@porcu2024matern.

@@

In view of these results, the parameter $\nu$ is suggested as a measure
of the smoothness of the PP index function by fitting a Gaussian process
prior with Matérn covariance on a dataset generated by random
evaluations of the index function. There exist several R packages,
such as `GpGp` [@guinness2021gpgp] or `ExaGeoStatR` [@abdulah2023large],
to fit the hyper-parameters of a GP covariance function on data, which
is usually done by maximum likelihood estimation. In this project, the
`GpGp` package is used.

```{=tex}
\begin{defn}
Let $\mathbf{A}=[A_1, \ldots, A_N] \in (\mathbb{R}^{p \times d})^N$ be d-dimensional projection bases, $\mathbf{y}=[f(XA_1),\ldots,f(XA_N)]$ be the corresponding PP index values, and $\mathbf{K}=[K_\theta(A_{i}-A_{j})]_{1\leq i,j\leq N}\in\mathbb{R}^{N\times N}$ be the Matérn covariance matrix evaluated at the input bases, where the vector $\theta$ contains all the parameters of the multivariate Matérn covariance function $K$ (smoothness, outputscale, lengthscales). The log-likelihood of the parameters $\theta$ is defined by 
\begin{equation}
\mathcal{L}(\theta)=\log p(\mathbf{y}\left|\mathbf{A},\theta\right.)=-\frac{1}{2}\mathbf{y}^{\top}(\mathbf{K}+\sigma^{2}\mathbf{I})^{-1}\mathbf{y}-\frac{1}{2}\mathrm{\log}(\det(\mathbf{K}+\sigma^{2}\mathbf{I}))-\frac{N}{2}\log(2\pi)\, \label{eq:gp_log_likelihood}
\end{equation}
where the nugget parameter $\sigma$ is the standard deviation of the intrinsic noise of the Gaussian process.
The optimal parameters (including smoothness) are obtained by maximum log-likelihood
\begin{equation}
\theta^* = \underset{\theta}{\max}\mathcal{L}(\theta)
\end{equation}
The resulting optimal smoothness parameter $\nu$ is chosen as our smoothness metric.
\end{defn}
```
The value of the optimal smoothness parameter $\nu>0$ can be naturally
interpreted as follows: the higher $\nu$, the smoother the index
function.

@@

Squintability relates to the notion of squint angle and is defined based on how the index value $f(XA)$ changes with respect to the projection distance over the course of the optimization.

```{=tex}
\begin{defn}[squint angle]\label{def:squint-angle}
Let $A$ and $B$ be two $d$-dimensional orthonormal matrices in $\mathbb{R}^p$. The squint angle $\theta$ between the subspace spanned by $A$ and $B$ is defined as the smallest principal angle between these subspaces: $\theta = \min_{i \in \{1, \cdots, d\}} \arccos(\tau_i)$, where $\tau_i$ are the singular values of the matrix $M = A^T B$ obtained from its singular value decomposition.
\end{defn}

\begin{defn}[projection distance]\label{def:proj-dist}
Let $A \in \mathbb{R}^{p \times d}$ be a $d$-dimensional orthonormal matrix, and let $A^*$ be the optimal matrix that achieves the maximum index value for a given data. The projection distance between $A$ 
and $A^*$, $r(A, A^*)$, is defined by
$r(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\,\rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}
```

@@

From @barnett1981interpreting and @laa_using_2020, a large squint angle
implies that the objective function value is close to optimal even when
the perfect view to see the structure is far away. A small squint angle
means that the PP index value improves substantially only when the
perfect view is close by. As such, low squintability implies rapid
improvement in the index value when near the perfect view. For PP, a
small squint angle is considered to be undesirable because it means that
the optimizer needs to be very close to be able to "see" the optimum.
Thus, it could be difficult for the optimizer to find the optimum.
<!--The
mathematical formulation of this intuition is proposed below:-->

@@

# The jellyfish search optimizer {#sec-JSO}

@@

To solve the optimization problem embedded in the PP guided tour, a
starting projection, an index function, the number of jellyfish, and the
maximum number of trials (tries) are provided as input. Then, the
current projection is evaluated by the index function. The projection is
then moved in a direction determined by a random factor, influenced by
how far along we are in the optimization process. Occasionally,
completely new directions may be taken like a jellyfish might with ocean
currents. A new projection is accepted if it is an improvement compared
to the current one, rejected otherwise. This process continues and
iteratively improves the projection, until the pre-specified maximum
number of trials is reached.

@@

**Output**: `optimized_projection`

@@

**return** the set of projections with the updated `current_best` as the
`optimized_projection`
:::

@@

# Assessing the optimizers {#sec-sim-deets}

This section explains the details of two simulation studies: (1) comparison between the existing creeping random search [@RJ-2021-105; @laa_using_2020] and JSO, and (2) examining the factors affecting the performance of JSO.

@@

The CRS is the main optimization routine currently used for the guided tour. Here the two optimizers are compared on the task of finding the pipe using the `holes` index in data with dimensions, $d = 6, 8, 10, 12$. Fifty simulated data sets are used for each dimension. JSO uses 100 jellyfish with a maximum of 100 tries, while the CRS allows a maximum of 1000 tries at each iteration before the algorithm terminates. These choices enable fair comparison between CRS and JSO, that conforms to how they are used in practice.
